Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
data

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpretted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%\hl{[I'm not familiar with bracket notation,
%  so these equations are confusing to me.  How widely used is this
%  notation in ecology?  Will others likely understand this?  Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be
estimated given that there is only a single observation available?
Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this
makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

A fourth problem with these models is that they do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models and it ignores the importance of space in
population regulation.
%, which
%restrict the utility of their since it is typically
%impossible assess the impacts of factors such as habitat
%fragmentation or climate change on population dynamics without
%considering spatial variation in abundance. Indeed many populations are
%regulated by spatial processes such as source-sink dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes.

In this paper, we focus on the model of
\citet[henceforth the DM model;]{dail_madsen:2011}
%developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frequentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

@@

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ time periods. A site is ideally, but not necessarily, a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat. In some cases, a site may not have a clear
biological significance, for instance in the case of randomly located
survey plots. The timeframe of the study is arbitrary, but the general
situation is one in which the time period is sufficiently long such
that abundance changes at each site. We will call
these time periods primary sampling periods to distinguish them from
secondary sampling periods, which are repeated surveys within a
primary period. In the case where no secondary sampling was conducted,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
j=1,\hdots,J$ is the count at site $i$ during secondary sampling
occasion $j$ within year $t$. These are the observed data, which
typically will less than the actual quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $\{N_{it}\}$ are the data.

@@

>>>>

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpretted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
%  so these equations are confusing to me.  How widely used is this
%  notation in ecology?  Will others likely understand this?  Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be
estimated given that there is only a single observation available?
Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this
makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

A fourth problem with these models is that they do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models and it ignores the importance of space in
population regulation.

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only the work by
\citet{kery_etal:2009} addressed several of these limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes.

In this paper, we focus on the model of
\citet[henceforth the DM model;]{dail_madsen:2011}
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frequentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

@@

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located
survey plots. The timeframe of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a one
month long breeding season, which could be surveyed once per year for
$T$ years. If the a site is surveyed on $J$ occassions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e. $J=1$,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If secondary sampling periods were used,
an extra dimension is added so tha twe have $n_{ijt}:
j=1,\hdots,J$. Typically, we $n_{it}$ will be lower than the actual
quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $n_{it} \equiv N_{it}$.

@@

\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}
